---
title: "scratchwork"
author: "Lillian Clark"
date: "11/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(survival)
library(survminer)
library(rstanarm)
```

```{r}
load("data/37692-0001-Data.rda")
load("data/37692-0002-Data.rda")
prison <- da37692.0001
state_only <- da37692.0002
```


```{r}
prison_clean <- prison %>%
  select(V0390, V0391, V0392, V0393, V0394)

colnames(prison_clean) <- c("x1", "year", "month", "week", "day")

prison_clean %>%
  count(x1)

prison %>%
  count(V1951) %>%
  arrange(desc(n))

prison %>%
  count(V0489) %>%
  arrange(desc(n))

#consider victimless crimes
#help_at_offense = V0478, victim_sex = V0489, 
#victim_race V0488 and before

prison %>%
  filter(V0489 %in% c("(1) 1 = Male", "(2) 2 = Female")) %>%
  count(RV0037)

prison %>%
  filter(V0488 %in% c("(1) 1 = Contained at least one valid response entry")) %>%
  count(RV0037)

prison %>%
  count(V0491)

prison %>%
  count(V0557)

prison %>%
  filter(V0491 %in% c("(1) 1 = Knew", "(2) 2 = Stranger")) %>%
  count(V0557)

prison_filt <- prison %>%
  select(current_age = RV0001, race = RV0003, citizen = RV0004,
         military = RV0008, controlling_offense = RV0036, controlling_offense_type = RV0037,
         education = RV0054, firearm_at_offense = RV0095,
         arrest_year = V0055Y, admit_year = V0056Y, held_by = V0071, sentenced = V0073, awaiting = V0075, 
         arrested_during_status = V0078, jail_time_served = V0390, jail_time_yr = V0391, jail_time_mo = V0392,
         jail_time_wk = V0393, jail_time_dy = V0394,
         victim_hispanic = V0481, victim_hispanic2 = V0544, victim_white = V0482,
         victim_white2 = V0545, victim_black = V0483, victim_black2 = V0546, 
         victim_native_american = V0484, victim_native_american2 = V0547, victim_asian = V0485, victim_asian2 = V0548,
         victim_hawaiian = V0486, victim_hawaiian2 = V0549, victim_race = V0488, victim_race2 = V0551,
         victim_sex = V0489, victim_age = V0490, victim_known = V0491, victim_offense = V0495,
         victim_injured = V0496, victim_died = V0497, state = V0772, sex = V1212, 
         alc_at_offense = V1267, drug_at_offense = V1326) %>%
  mutate(
    across(c("jail_time_yr", "jail_time_mo", "jail_time_wk", "jail_time_dy"), 
                ~ifelse(.x < 0, as.numeric(NA), .x)),
    jail_time_yr = ifelse(jail_time_yr > 40, as.numeric(NA), jail_time_yr),
    time_estimate = admit_year - arrest_year)
  rowwise() %>%
  mutate(
    jail_time = case_when(
      is.na(jail_time_yr) & is.na(jail_time_mo) & is.na(jail_time_wk) & is.na(jail_time_dy) ~ as.numeric(NA),
      TRUE ~ round(sum(c(jail_time_yr * 365, jail_time_mo * 365 / 12, jail_time_wk * 7, jail_time_dy), 
                       na.rm = TRUE), 0))) %>%
    ungroup()

prison_filt %>%
  count(victim_hispanic)

attributes(prison_filt)$variable.labels <- NULL

prison_mod <- prison_filt %>%
  select(current_age, race, sex, citizen, military, education, jail_time_served, jail_time,
         arrest_year, admit_year, held_by, sentenced, controlling_offense, controlling_offense_type,
         arrested_during_status, firearm_at_offense, alc_at_offense, drug_at_offense, 
         victim_hispanic, victim_hispanic2, victim_white, victim_white2, victim_black, victim_black2, 
         victim_native_american, victim_native_american2, victim_asian, victim_asian2,
         victim_hawaiian, victim_hawaiian2, victim_race, victim_race2,
         victim_sex, victim_age, victim_known, victim_offense) %>%
  mutate(across(c("race", "citizen", "military", "controlling_offense", "controlling_offense_type",
                  "education", "firearm_at_offense", "held_by", "sentenced", "arrested_during_status",
                  "jail_time_served", "sex", "alc_at_offense", "drug_at_offense", starts_with("victim")), 
                ~ str_remove(as.character(.x), "\\(\\d+\\) \\d+ = ")),
         across(c("arrest_year", "admit_year"), ~ifelse(.x == 999999, as.numeric(NA), .x)),
         across(c("alc_at_offense", "drug_at_offense", "sentenced", starts_with("victim")), 
                ~ifelse(.x %in% c("(-1) -1 = Don't Know", "(-2) -2 = Refusal"), as.character(NA), .x)),
         race = str_remove(race, " \\(NH\\)"),
         race = case_when(
           race == "Uncategorized - Missing" ~ as.character(NA), 
           race == "2+ Races" ~ "Other",
           TRUE ~ race),
         citizen = ifelse(citizen == "Missing", as.character(NA), citizen),
         military = ifelse(military == "(98) 98 = DK/REF", as.character(NA), military),
         controlling_offense_type = 
           ifelse(controlling_offense_type == "DK/REF", as.character(NA), controlling_offense_type),
         education = ifelse(education == "Missing", as.character(NA), education),
         firearm_at_offense = 
           ifelse(firearm_at_offense %in% c("DK/REF", "Missing (in-universe)"), as.character(NA), firearm_at_offense),
         held_by = case_when(
           held_by == "State correctional authorities such as the state department of corrections" ~ "State",
           held_by %in% c("Federal Bureau of Prisons", "U.S. Marshals Service") ~ "Federal",
           held_by == "Local correctional authorities such as local jails or detention centers" ~ "Local",
           held_by == "U.S. Immigration and Customs Enforcement" ~ "ICE",
           held_by == "Some Other Authority" ~ "Other",
           TRUE ~ as.character(NA)),
         arrested_during_status = case_when(
           arrested_during_status == "None of These (No Parole, Probation, or Escape)" ~ "None",
           arrested_during_status == "Probation, including Shock Probation and Split Sentences" ~ "Probation",
           arrested_during_status == "Parole or Post-Release Supervision after serving time" ~ "Parole",
           arrested_during_status %in% c("(-2) -2 = Refusal", "(-1) -1 = Don't Know") ~ as.character(NA),
           TRUE ~ arrested_during_status),
         max_time_est = admit_year - arrest_year,
         max_time_est = ifelse(max_time_est < 0, as.numeric(NA), max_time_est),
         jail_time_served = case_when(
           jail_time_served == "Specify Jail Time" ~ "Yes",
           jail_time_served == "No Time in Jail" ~ "No",
           TRUE ~ as.character(NA)),
         jail_time = ifelse(jail_time_served == "No", 0, jail_time),
         victim_race = case_when(
           victim_race == "Contained a don't know response" ~ "Unknown",
           victim_race == "Contained at least one valid response entry" ~ 
             str_remove_all(str_remove_all(paste0(victim_white, ",", victim_black, ",", victim_native_american, 
                    ",", victim_asian, ",", victim_hawaiian), "Blank"), ","),
           TRUE ~ as.character(NA)),
         victim_race = case_when(
         victim_race %in% c("White" , "Unknown") ~ victim_race,
         victim_race == "Black or African American" ~ "Black",
         victim_race == "American Indian or Alaska Native" ~ "American Indian/Alaska Native",
         victim_race %in% c("Native Hawaiian or Other Pacific Islander", 
                            "Asian") ~ "Asian/Native Hawaiian/Other Pacific Islander",
         is.na(victim_race) ~ victim_race,
         TRUE ~ "Multiple"),
         victim_race2 = case_when(
           victim_race2 == "Contained a don't know response" ~ "Unknown",
           victim_race2 == "Contained at least one valid response entry" ~ 
             str_remove_all(str_remove_all(paste0(victim_white2, ",", victim_black2, ",", victim_native_american2, 
                    ",", victim_asian2, ",", victim_hawaiian2), "Blank"), ","),
           TRUE ~ as.character(NA)),
         victim_race2 = case_when(
         victim_race2 %in% c("White" , "Unknown") ~ victim_race2,
         victim_race2 == "Black or African American" ~ "Black",
         victim_race2 == "American Indian or Alaska Native" ~ "American Indian/Alaska Native",
         victim_race2 %in% c("Native Hawaiian or Other Pacific Islander", 
                            "Asian") ~ "Asian/Native Hawaiian/Other Pacific Islander",
         is.na(victim_race2) ~ victim_race2,
         TRUE ~ "Multiple"),
         victim_race = ifelse(is.na(victim_race), victim_race2, victim_race),
         victim_hispanic = ifelse(is.na(victim_hispanic), victim_hispanic2, victim_hispanic)) %>%
  select(-victim_hispanic2, -victim_white, -victim_white2, -victim_black, -victim_black2, 
         -victim_native_american, -victim_native_american2, -victim_asian, -victim_asian2,
         -victim_hawaiian, -victim_hawaiian2, -victim_race2)

prison_mod %>%
  count(sentenced)

prison_mod %<>%
  mutate(censored = case_when(
    sentenced == "No" ~ 0,
    sentenced == "Yes" ~ 1,
    TRUE ~ as.numeric(NA)))

# what to do about these
prison_mod %>%
  filter(max_time_est > 0 & jail_time / 365 > max_time_est) %>%
  select(jail_time, max_time_est, admit_year, arrest_year)

prison_mod %>%
  filter(is.na(jail_time))

prison_mod %>%
  count(race) %>%
  arrange(desc(n))

prison_mod %>%
  count(victim_died) %>%
  arrange(desc(n))


prison_mod %>%
  ggplot(aes(x = jail_time)) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 2500))

prison_filt %>%
  count(jail_time) %>%
  arrange(desc(jail_time))
```

```{r}
data <- prison_mod %>%
  select(!starts_with("victim"))

data_v <- prison_mod %>%
  filter(controlling_offense_type == "Violent")



prison_mod %>%
  count(controlling_offense, is.na(victim_sex)) %>%
  arrange(controlling_offense)

prison_mod %>%
  count(controlling_offense_type, controlling_offense)

data2 <- prison_mod %>%
  filter()
```

```{r}
survdata <- data

ggsurvplot(survfit(Surv(jail_time, censored) ~ race, data = survdata),
     legend.title = "",
     conf.int = F, censor = F,
     xlim = c(0, 2500),
     ggtheme = theme_minimal())

ggsurvplot(survfit(Surv(jail_time, censored) ~ victim_race, data = data_v),
     legend.title = "",
     conf.int = F, censor = F,
     xlim = c(0, 2500),
     ggtheme = theme_minimal())



logrank_test <- survdiff(Surv(yos, retired) ~ onedone01, data = surv_data)
logrank_p <- round(glance(logrank_test)$p.value[[1]], 4)
tidy_logrank <- tidy(logrank_test) %>%
  mutate(calc_oe = ((obs - exp) ^ 2) / exp,
         calc_oez = ((obs - exp) ^ 2 ) / diag(logrank_test$var),
         onedone01 = ifelse(onedone01 == 1, "One-and-Done", "Longer College Stay"))

```

race, 

maybe model time until trial??? or sentencing?? so whether or not still awaiting trial = censoring, time in jail + any additional prison time = days, lots of demographic factors as covariates and potentially other values

arrest date
admission to prison date

censoring: sentenced

V0098
V0412

how long people wait for trial
possible focuses: were they homeless (V0961), race, 

V1040 on is about child contact/visitation
V1172 on whether various fam members incarcerated
mental health section!!

V1217 seen healthcare provider

what contributes to how much time people spend in jail pre-trial

outcome = time in jail + time in prison pre-sentencing
censoring = have you been sentenced (no = censored)
look at attributes law enforcement and DA would know about -- racial groups, crime type, firearm, victim type

