---
title: 'Waiting for Trial: A Case Study of Detention Times Prior to Sentencing'
author: "Lillian Clark"
date: "11/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(survminer)
library(rstanarm)
```

```{r load-data, message=FALSE}
prison <- read_csv("data/prison-clean.csv")

# data for first analysis
data <- prison %>%
  select(!starts_with("victim"))

# data for second analysis
data_v <- prison %>%
  filter(controlling_offense_type == "Violent")
```

Motivation:
Pre-trial detention is significant and destabilizing. People held in limbo by the criminal legal system for years. Some data available for scraping from jails but lots of problems with transparency here. ICPSR Survey of Prison Inmates 2016 allows a look at a large population across the U.S. and how long they reported they were detained in jails. The majority have been sentenced, although some reported that they had not been sentenced -- for these individuals, the time spent waiting for trial, or pre-sentencing, is censored. This data is restricted to individuals spending time in prison, so will exclude a majority of those with dismissed or not guilty charges who were also held pretrial. However, it still provides an important window into the punishment people face in America before they are even tried, the wait for justice.

Research questions:

1) When controlling for aspects of the main offense they are prosecuted for, do individuals incarcerated in United States prisons in different racial and ethnic groups wait longer before sentencing? Are there other factors that contribute to longer time spent in pre-trial detention?

2) Does time detained pre-trial have any relationship with characteristics of the victims of "violent" crimes, such as sex, race, age, and whether they are known by the detained individual?

TO DO:

- what to do with people whose jail time exceeded their estimated time with admit - arrest?

### Introduction

Outline:

- In 2019, 23.3% of incarcerated people in the United States were legally innocent -- they had not been tried on their charges in a court of law.^[World Prison Brief https://www.prisonstudies.org/country/united-states-america]. The Prison Policy Initiative that the number of people in jail pre-trial has nearly quadrupled since the 1980s.^[https://www.prisonpolicy.org/research/pretrial_detention/] Pretrial detention is severely destabilizing for incarcerated individuals and their families. Even a few days in jail can cause people to lose their jobs, their homes, or custody of their children. Studies have shown that people who are detained pretrial are more likely to be convicted and more likely to receive harsher sentences than those who go free. Pretrial detention is also associated with significantly higher recidivism rates. 
- Widespread use of money bail has led to the criminalization of poverty, where individuals who can afford to pay a bail bondsman's fee or to tie up thousands of their own dollars in the court system for months or years can go free and individuals without those funds sit in jail. This can force innocent individuals who can't pay to claim guilt and accept a plea deal simply to get out faster. In the U.S., there are also clear racial disparities among pretrial detainees. Young Black men are roughly 50% more likely to be detained than white individuals.^[Prison Policy Initiative]. Some people are held in limbo by the criminal legal system for amounts of time comparable to what they would serve if they were convicted. In 2017, The New York Times profiled Kharon Davis, a man who spent at least ten years in county jail awaiting trial. They wrote, "Though he has not been found guilty, Mr. Davis has already served half of the minimum sentence for murder."^[https://www.nytimes.com/2017/09/19/us/alabama-kharon-davis-speedy.html]
- "Non-white defendants receive bail amounts that are twice as high as bail set for white defendants and are also less likely to be able to afford it. Black and American Indian women are also overrepresented in US pre-trial facilities."
- Not everyone held in jail is awaiting trial -- there are also individuals with relatively minor convictions (like parole violations or misdemeanors) that serve their sentences there (usually for less than a year). The proportion of pretrial detainees in jail populations has increased from 53% in 1970 to 64% in 2015.^[Vera Brief https://www.vera.org/downloads/publications/Justice-Denied-Evidence-Brief.pdf] Some data is available for scraping from local jails but there are significant problems with transparency, inaccuracies or data entry issues. These problems can make it hard to do things like parse our which jail inhabitants have been convicted and which have not. It makes studying pretrial detention and inequities harder.^[The Transparency of Jail Data, https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3741638]
- The ICPSR 2016 Survey of Prison Inmates contains data from 364 prisons across the United States. 24,848 incarcerated individuals participated, 81% of which were held at the time in state prisons and the other 19% in federal prisons. Part of the survey asked respondents how long they had been detained in jail.
- Error involved in any kind of survey data: jail time is self-reported and not verified from a second source. Also, the survey question doesn't ask participants to specify whether this jail time was pre-detention. Using jail time as length of pre-trial detention is not a perfect proxy, but we have reason to believe it is a good one (at least given the way the system is supposed to work), since those serving their sentences for minor convictions in jails should not have been surveyed and split jail-prison sentences are not the norm -- generally, individuals in prison, if they were detained pretrial, did so in jails, and were transferred to prisons after their convictions. 
- 391 individuals, or 1.6% of all observations in the data, reported that they had not yet been sentenced. These pretrial wait times are right-censored, but the group is very small compared to the overall survey participants.
- This data is restricted to individuals in prison, so will exclude those with dismissed or not guilty charges who were also held pretrial. However, it still provides an important window into the punishment people face in America before they are even tried (the wait for justice).


